---
title: 'Titanic: Machine Learning from Disaster'
author: "Ian Santillan"
date: "May 19, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Titanic: Machine Learning from Disaster

Predicting the survival rate on the Titanic

The sinking of the RMS Titanic is one of the most infamous shipwrecks in history.  On April 15, 1912, during her maiden voyage, the Titanic sank after colliding with an iceberg, killing 1502 out of 2224 passengers and crew. This sensational tragedy shocked the international community and led to better safety regulations for ships.

One of the reasons that the shipwreck led to such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others, such as women, children, and the upper-class.

In this analysis, using Binary Classifcation and R, we will predict what sorts of people were likely to survive. 

```{r Titanic}
# set working directory
setwd("C:/_scripts/Titanic")

# get train data
titanic.train <- read.csv(file = "train.csv", stringsAsFactors = FALSE, header = TRUE)

# get test data
titanic.test <- read.csv(file = "test.csv", stringsAsFactors = FALSE, header = TRUE)

# get median of age
median(titanic.train$Age)
median(titanic.train$Age, na.rm = TRUE)
median(titanic.test$Age, na.rm = TRUE)


```

## Create a unique identifier to combine the 2 datasets

Create a new column that doesn't exist

IsTrainSet

```{r}

titanic.train$IsTrainSet <- TRUE
tail(titanic.train$IsTrainSet)
titanic.test$IsTrainSet <- FALSE
tail(titanic.test$IsTrainSet)

ncol(titanic.train)
ncol(titanic.test)

# notice how the test set is missing the 'survive' column


```
## Combine the 2 data sets together to clean them both at once

To combine 2 datasets together, they need to be identical; meaning, the same column names, the same number of columns

```{r}
# check column names
names(titanic.test)
names(titanic.train)

# create a new column in the test set and fill it with NA
titanic.test$Survived <- NA

names(titanic.test)
names(titanic.train)

# verify number of columns in both datasets
ncol(titanic.train)
ncol(titanic.test)

titanic.full <- rbind(titanic.train, titanic.test)

# confirm if any rows are missed
418 + 891

tail(titanic.full)

table(titanic.full$IsTrainSet)

```
## clean the data

titanic.full$Embarked has 2 missing values.

titanic.full$age has lots of missing values.

titanic.full$Fare has 1 missing values.
```{r}

## titanic.full$Embarked

# notice the 2 values that missing the embarked
table(titanic.full$Embarked)

# S is the mode, use this to replace the missing value
titanic.full[titanic.full$Embarked=='', "Embarked"] <- 'S'

# verify that there is no more missing value
table(titanic.full$Embarked)


## titanic.full$Age

# check all the missing values for titanic.full$Age
table(is.na(titanic.full$Age))

# there are 263 missing values for titanic.full$Age. Replace the missing with the median

# get the median for titanic.full$Age
age.median <- median(titanic.full$Age, na.rm = TRUE)

# replace the missing value with the median
titanic.full[is.na(titanic.full$Age), "Age"] <- age.median

#verify that there is no more missing value
table(is.na(titanic.full$Age))


## titanic.full$Fare
table(is.na(titanic.full$Fare))

# get the median for titanic.full$Fare
fare.median <- median(titanic.full$Fare, na.rm = TRUE)

# replace the missing value with the median
titanic.full[is.na(titanic.full$Fare), "Fare"] <- fare.median

#verify that there is no more missing value
table(is.na(titanic.full$Fare))
```

## now that our data is clean, we are now ready to build a predctive model

Check the dataset structure first to ensure that the data types are correct.

We need to cast Pclass, Sex and Embarked as factors.

We'll need to split our data out into training and test

```{r}
# check str of dataset
str(titanic.full)

# categorical casting (except for Survived)
titanic.full$Pclass <- as.factor(titanic.full$Pclass)
titanic.full$Sex <- as.factor(titanic.full$Sex)
titanic.full$Embarked <- as.factor(titanic.full$Embarked)

# train set
titanic.train <- titanic.full[titanic.full$IsTrainSet ==  TRUE,]

# test set
titanic.test <- titanic.full[titanic.full$IsTrainSet !=  TRUE,]


# check str of dataset
str(titanic.full)

class(titanic.full$Survived)

# cast Survive as categorical
titanic.full$Survived <- as.factor(titanic.full$Survived)

# verify new class for full$Survived has been set
class(titanic.full$Survived)

# check to see how many survived
table(titanic.full$Survived)
```
## define which columns you want to use to build your model


```{r}
# check the structure to see which columns we need to use
str(titanic.full)

survived.equation <- "Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked"

survived.forumla <- as.formula(survived.equation)

```


## we can now create our predictive model

```{r}
# load the randomForest package
install.packages("randomForest")
library(randomForest)

titanic.model <- randomForest(formula = survived.forumla, data = titanic.train, ntree = 500, mtry = 3, nodesize = 0.01 * nrow(titanic.test))

```

## now let's apply this predictive model

define what features are being used

it's important that we define the features because we want to use the PassengerID

```{r}

features.equation <- "Pclass + Sex + Age + SibSp + Parch + Fare + Embarked"

Survived <- predict(titanic.model, newdata = titanic.test)

```

## output

```{r}

PassengerID <- titanic.test$PassengerId
output.df <- as.data.frame(PassengerID)
output.df$Survived <- Survived

# check the tail of the dataframe
tail(output.df)

# output to csv
write.csv(output.df, file='titanic_test_survival.csv', row.names = FALSE)

```

## future items
parameter tuning
cross validation
70/30 split
build a regression model
build a predictive model to predict missing values of Age/Fare instead of using median